---
title: "USD/PHP FX Risk Analysis | Items 1–6"
output:
  html_document:
    df_print: kable
    toc: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 6)
library(readr)
library(dplyr)
library(lubridate)
library(knitr)
library(e1071)
library(ggplot2)
library(ggrepel)
library(tidyr)

# --- ECONOMIST & QVRS THEME ---
theme_economist_quant <- function() {
  theme_minimal() +
    theme(
      text = element_text(family = "sans"),
      plot.title = element_text(face = "bold", size = 16, color = "#0D0D0D"),
      plot.subtitle = element_text(size = 11, color = "#595959", margin = margin(b = 10)),
      plot.caption = element_text(size = 8, color = "#B3B3B3", hjust = 0, margin = margin(t = 10)),
      panel.grid.major.y = element_line(color = "#EBEBEB", linewidth = 0.3),
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      axis.title = element_text(size = 10, face = "bold", color = "#333333"),
      axis.ticks = element_blank(),
      legend.position = "bottom",
      legend.title = element_blank()
    )
}

econ_blue <- "#141F52"
econ_red  <- "#E3120B"
econ_grey <- "#595959"
econ_gold <- "#D4A017"

# Helper for Levene's Test (Manual Implementation — median-centred)
levene_test_manual <- function(group1, group2) {
  med1 <- median(group1)
  med2 <- median(group2)
  z1 <- abs(group1 - med1)
  z2 <- abs(group2 - med2)
  t_res <- t.test(z1, z2, var.equal = TRUE)
  return(list(statistic = t_res$statistic, p_value = t_res$p.value))
}
```

# Section 1: Data Load and Integrity Check

Data cut-off: **27 February 2026**, per case guidelines. This analysis uses the unified master dataset containing only raw published trading days from the BAP (no imputed values). The 2021-05-20 Open typo (478.855 → 47.855) has been corrected.

```{r data_load}
df_raw <- read_csv("data/unified/unified_bap_master.csv") %>%
  mutate(Date = as.Date(Date)) %>%
  # Enforce case cut-off: 27 Feb 2026
  filter(Date <= as.Date("2026-02-27")) %>%
  arrange(Date)

integrity_check <- data.frame(
  Metric = c("Total Trading Rows", "Start Date", "End Date", "Rows with High < Low", "Rows with Close outside High-Low range"),
  Value = c(
    as.character(nrow(df_raw)),
    as.character(min(df_raw$Date)),
    as.character(max(df_raw$Date)),
    as.character(nrow(filter(df_raw, High < Low))),
    as.character(nrow(filter(df_raw, Close > High | Close < Low)))
  )
)
kable(integrity_check, caption = "Data Integrity Check")
```

```{r price_history_viz}
# --- PLOT 0: The Full Picture ---
# Shade the three short windows to contextualise their size vs history
shade_df <- data.frame(
  xmin = as.Date(c(tail(df_raw$Date, 60)[1], tail(df_raw$Date, 40)[1], tail(df_raw$Date, 20)[1])),
  xmax = max(df_raw$Date),
  label = c("60-day", "40-day", "20-day"),
  y    = c(54.5, 55.8, 57.1)
)

ggplot(df_raw, aes(x = Date, y = Close)) +
  geom_rect(data = shade_df,
            aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = label),
            inherit.aes = FALSE, alpha = 0.12) +
  scale_fill_manual(values = c("60-day" = econ_gold, "40-day" = econ_grey, "20-day" = econ_red),
                    name = "Analysis window") +
  geom_line(color = econ_blue, linewidth = 0.6) +
  annotate("text", x = as.Date("2020-04-01"), y = 51.2,
           label = "COVID-19\nshock", size = 3, color = econ_grey, hjust = 0) +
  annotate("text", x = as.Date("2022-06-01"), y = 55.8,
           label = "2022 PHP\ndepreciation", size = 3, color = econ_grey, hjust = 0) +
  labs(title = "Seven years of USD/PHP",
       subtitle = "Daily closing rate, 2 Jan 2019 – 27 Feb 2026 (BAP data). Shaded regions show the three short analysis windows.",
       x = NULL, y = "PHP per USD",
       caption = "Source: BAP Historical Data. Cut-off: 27 Feb 2026.") +
  theme_economist_quant()
```

# Section 2: Data Cleaning

```{r data_cleaning}
# Correct High/Low bounds to encompass Open and Close (robust integrity)
df <- df_raw %>%
  mutate(
    ActualHigh = pmax(Open, High, Low, Close),
    ActualLow  = pmin(Open, High, Low, Close),
    High = ActualHigh,
    Low  = ActualLow
  ) %>%
  select(-ActualHigh, -ActualLow)

# Derive return columns
df <- df %>%
  mutate(
    dod_cents    = (Close - lag(Close)) * 100,
    dod_pct      = (Close - lag(Close)) / lag(Close) * 100,
    intraday_cents = (High - Low) * 100,
    intraday_pct   = (High - Low) / Low * 100
  )

# Drop the first row (no prior close → NA returns)
df_returns <- df %>% filter(!is.na(dod_cents))

cleaning_stats <- data.frame(
  Metric = c("Modified High/Low Bounds", "Derived Columns", "Remaining Trading Rows (DoD series)"),
  Result = c(
    as.character(nrow(filter(df_raw, High < Low | Close > High | Close < Low | Open > High | Open < Low))),
    "DoD cents, DoD %, Intraday cents, Intraday %",
    as.character(nrow(df_returns))
  )
)
kable(cleaning_stats, caption = "Data Cleaning Results Summary")
```

# Section 3: Window Definitions

```{r windows}
# Return series windows (for DoD / Items 1, 4, 5, 6)
windows_returns <- list(
  Full = df_returns,
  w60  = tail(df_returns, 60),
  w40  = tail(df_returns, 40),
  w20  = tail(df_returns, 20)
)

# Intraday series windows (for Item 2; no lag dependency so use df)
windows_intraday <- list(
  Full = df,
  w60  = tail(df, 60),
  w40  = tail(df, 40),
  w20  = tail(df, 20)
)

window_summary <- data.frame(
  Window     = names(windows_returns),
  Start_Date = sapply(windows_returns, function(x) as.character(min(x$Date))),
  End_Date   = sapply(windows_returns, function(x) as.character(max(x$Date))),
  n          = sapply(windows_returns, nrow)
)
kable(window_summary, caption = "Analysis Windows (Return Series, cut-off 27 Feb 2026)")
```

# Section 4: Item 1 — Close-to-Close VaR (95%, one-tail)

VaR is expressed as **loss magnitude** (positive number). Using the 5th-percentile of the return distribution — i.e., we expect a loss *at least* this large on 5% of trading days.

```{r item1_tables}
# Use factor for ordered output in tables and plots
window_order <- c("Full", "w60", "w40", "w20")

item1_var <- bind_rows(lapply(windows_returns, function(w) {
  data.frame(
    n         = nrow(w),
    VaR_Pct   = abs(quantile(w$dod_pct,   0.05, type = 7)),
    VaR_Cents = abs(quantile(w$dod_cents,  0.05, type = 7))
  )
}), .id = "Window") %>%
  mutate(Window = factor(Window, levels = window_order)) %>%
  arrange(Window)

kable(item1_var, digits = 4,
      caption = "Item 1: Close-to-Close VaR at 95% Confidence (Loss Magnitude)")

# Outliers: observations outside the 0.1% / 99.9% bounds of the full history
lb_dod <- quantile(df_returns$dod_pct, 0.001, type = 7)
ub_dod <- quantile(df_returns$dod_pct, 0.999, type = 7)

outliers_1 <- df_returns %>%
  filter(dod_pct < lb_dod | dod_pct > ub_dod) %>%
  select(Date, Close, dod_pct, dod_cents) %>%
  arrange(Date)

kable(outliers_1, digits = 4,
      caption = paste0("Item 1: Outliers — DoD % outside [",
                       round(lb_dod, 4), "%, ", round(ub_dod, 4), "%] (0.1%/99.9% bounds)"))
```

```{r item1_viz}
# --- PLOT 1: The Long Tail of Risk ---
full_var_pct <- item1_var$VaR_Pct[item1_var$Window == "Full"]

ggplot(df_returns, aes(x = dod_pct)) +
  # Empirical density
  geom_density(fill = econ_blue, alpha = 0.20, color = econ_blue, linewidth = 0.7) +
  # Normal reference curve
  stat_function(fun = dnorm,
                args = list(mean = mean(df_returns$dod_pct), sd = sd(df_returns$dod_pct)),
                linetype = "dashed", color = econ_grey, linewidth = 0.6) +
  # VaR line (left tail — negative, because VaR magnitude is positive)
  geom_vline(xintercept = -full_var_pct, color = econ_red, linewidth = 1) +
  geom_label(aes(x = -full_var_pct, y = 0.30,
                 label = paste0("VaR 95%: ", round(full_var_pct, 3), "%")),
             size = 3.2, color = econ_red, fill = "white", label.size = 0, vjust = 0) +
  # Outlier markers
  geom_rug(data = outliers_1, aes(x = dod_pct), color = econ_red, linewidth = 0.8, alpha = 0.8) +
  geom_text_repel(data = outliers_1,
                  aes(x = dod_pct, y = 0.02,
                      label = format(Date, "%b %Y")),
                  size = 2.8, color = econ_red, nudge_y = 0.05, segment.size = 0.3) +
  labs(title = "The long tail of risk",
       subtitle = "Density of USD/PHP daily % changes vs. Normal (dashed). Red rug marks = extreme outliers (outside 99.9%).",
       x = "Daily Close-to-Close Change (%)", y = "Density",
       caption = "Source: BAP. Full history 2019–2026. VaR = 5th percentile of empirical distribution (type 7).") +
  theme_economist_quant()
```

# Section 5: Item 2 — Intraday (High-to-Low) VaR (95%, one-tail)

Intraday range is always non-negative, so the one-tail 95% test uses the **95th percentile** (upper tail).

```{r item2}
item2_var <- bind_rows(lapply(windows_intraday, function(w) {
  data.frame(
    n         = nrow(w),
    VaR_Cents = quantile(w$intraday_cents, 0.95, type = 7),
    VaR_Pct   = quantile(w$intraday_pct,   0.95, type = 7)
  )
}), .id = "Window") %>%
  mutate(Window = factor(Window, levels = window_order)) %>%
  arrange(Window)

kable(item2_var, digits = 4,
      caption = "Item 2: Intraday (High–Low) VaR at 95% Confidence (95th percentile of range)")

# Outliers: above 99.9th percentile (range is one-sided positive)
ub_intraday <- quantile(df$intraday_cents, 0.999, type = 7)
outliers_2 <- df %>%
  filter(intraday_cents > ub_intraday) %>%
  select(Date, High, Low, intraday_cents, intraday_pct) %>%
  arrange(Date)

kable(outliers_2, digits = 4,
      caption = paste0("Item 2: Outliers — Intraday cents above ",
                       round(ub_intraday, 4), " cents (99.9th percentile)"))
```

```{r item2_viz}
# --- PLOT 1b: Intraday volatility scatter ---
ggplot(df, aes(x = Date, y = intraday_cents)) +
  geom_point(color = econ_blue, alpha = 0.35, size = 0.9) +
  geom_smooth(method = "loess", span = 0.15, se = FALSE, color = econ_red, linewidth = 0.8) +
  geom_hline(yintercept = item2_var$VaR_Cents[item2_var$Window == "Full"],
             linetype = "dashed", color = econ_grey, linewidth = 0.6) +
  annotate("text", x = as.Date("2019-06-01"),
           y = item2_var$VaR_Cents[item2_var$Window == "Full"] + 1.5,
           label = paste0("VaR 95% (full): ",
                          round(item2_var$VaR_Cents[item2_var$Window == "Full"], 2), " cts"),
           size = 3, color = econ_grey, hjust = 0) +
  geom_point(data = outliers_2, aes(x = Date, y = intraday_cents),
             color = econ_red, size = 3) +
  geom_text_repel(data = outliers_2, aes(x = Date, y = intraday_cents,
                  label = format(Date, "%b %Y")),
                  size = 2.8, color = econ_red, nudge_y = 2) +
  labs(title = "Wide days — intraday volatility over time",
       subtitle = "Each point = one trading day's High–Low range (cents). LOESS trend in red. Dashed = full-history VaR 95%.",
       x = NULL, y = "Intraday Range (cents)",
       caption = "Source: BAP. Red dots = outliers above 99.9th percentile.") +
  theme_economist_quant()
```

# Section 6: Item 3 — Closing vs. Intraday: The Volatility Gap

```{r item3_table}
gap_table <- item1_var %>%
  select(Window, dod_var = VaR_Cents) %>%
  left_join(item2_var %>% select(Window, intra_var = VaR_Cents), by = "Window") %>%
  mutate(
    Gap_Cents = intra_var - dod_var,
    Gap_Pct_of_Intra = Gap_Cents / intra_var * 100
  )
kable(gap_table, digits = 4,
      caption = "Item 3: VaR Gap (Intraday − DoD). Positive gap = intraday captures more risk than close-to-close.")
```

```{r item3_viz}
# --- PLOT 2: Mind the Gap ---
# Show full history (not just 250 days) to avoid cherry-picking the window
ggplot(df_returns, aes(x = Date)) +
  geom_hline(yintercept = 0, color = econ_grey, linewidth = 0.3) +
  geom_line(aes(y = intraday_pct,   color = "Intraday Range (H-L)"), linewidth = 0.45, alpha = 0.7) +
  geom_line(aes(y = dod_pct,        color = "Close-to-Close Change"), linewidth = 0.45, alpha = 0.8) +
  scale_color_manual(values = c("Intraday Range (H-L)" = econ_blue,
                                "Close-to-Close Change"  = econ_red)) +
  labs(title = "Mind the gap",
       subtitle = "Intraday H-L range (%) vs signed close-to-close change (%), full history 2019–2026.",
       x = NULL, y = "Change / Range (%)",
       caption = paste0("Full history. Intraday VaR 95% = ",
                        round(item2_var$VaR_Pct[item2_var$Window == "Full"], 4), "% vs DoD VaR 95% = ",
                        round(item1_var$VaR_Pct[item1_var$Window == "Full"], 4),
                        "%. Note how intraday range often dwarfs the net daily move.")) +
  theme_economist_quant()
```

# Section 7: Item 4 — Coherent Risk Measures: CVaR & Winsorization

```{r item4}
# CVaR (Expected Shortfall) — mean of returns below the 5th percentile
threshold_pct   <- -item1_var$VaR_Pct[item1_var$Window == "Full"]
threshold_cents <- -item1_var$VaR_Cents[item1_var$Window == "Full"]

cvar_pct   <- abs(mean(df_returns$dod_pct[df_returns$dod_pct     <= threshold_pct]))
cvar_cents <- abs(mean(df_returns$dod_cents[df_returns$dod_cents <= threshold_cents]))

# Winsorization at 0.1% / 99.9% (caps extreme values at boundary rather than removing them)
lb_w <- quantile(df_returns$dod_pct,   0.001, type = 7)
ub_w <- quantile(df_returns$dod_pct,   0.999, type = 7)
lb_c <- quantile(df_returns$dod_cents, 0.001, type = 7)
ub_c <- quantile(df_returns$dod_cents, 0.999, type = 7)

df_winsor <- df_returns %>%
  mutate(
    dod_pct_w   = pmin(pmax(dod_pct,   lb_w), ub_w),
    dod_cents_w = pmin(pmax(dod_cents, lb_c), ub_c)
  )

var_winsor_pct   <- abs(quantile(df_winsor$dod_pct_w,   0.05, type = 7))
var_winsor_cents <- abs(quantile(df_winsor$dod_cents_w, 0.05, type = 7))

raw_var_pct   <- item1_var$VaR_Pct[item1_var$Window == "Full"]
raw_var_cents <- item1_var$VaR_Cents[item1_var$Window == "Full"]

item4_table <- data.frame(
  Metric = c(
    "Raw VaR (95%)",
    "CVaR / Expected Shortfall (95%)",
    "Winsorization Lower Bound (0.1%)",
    "Winsorization Upper Bound (99.9%)",
    "VaR after Winsorization (95%)",
    "VaR Reduction from Winsorization"
  ),
  Percentage = c(raw_var_pct, cvar_pct, lb_w, ub_w, var_winsor_pct,
                 raw_var_pct - var_winsor_pct),
  Cents      = c(raw_var_cents, cvar_cents, NA, NA, var_winsor_cents,
                 raw_var_cents - var_winsor_cents)
)
kable(item4_table, digits = 4, caption = "Item 4: CVaR and Winsorization Results (Full History)")
```

```{r item4_viz}
# --- PLOT 3a: CVaR vs VaR tail illustration ---
tail_returns <- df_returns %>% filter(dod_pct <= threshold_pct)
cvar_line <- mean(tail_returns$dod_pct)

ggplot(df_returns, aes(x = dod_pct)) +
  geom_density(fill = econ_blue, alpha = 0.18, color = econ_blue, linewidth = 0.7) +
  # Shade the tail (ES zone)
  geom_density(data = tail_returns, aes(x = dod_pct),
               fill = econ_red, alpha = 0.4, color = NA) +
  geom_vline(xintercept = threshold_pct, color = econ_red, linewidth = 1) +
  geom_vline(xintercept = cvar_line, color = econ_gold, linewidth = 1, linetype = "dashed") +
  annotate("text", x = threshold_pct, y = 0.35,
           label = paste0("VaR 95%\n", round(-threshold_pct, 3), "%"),
           size = 3, color = econ_red, hjust = 1.1) +
  annotate("text", x = cvar_line, y = 0.35,
           label = paste0("CVaR (ES)\n", round(cvar_pct, 3), "%"),
           size = 3, color = econ_gold, hjust = 1.1) +
  labs(title = "Beyond the threshold: Expected Shortfall",
       subtitle = "Red shaded region = losses beyond VaR 95%. CVaR (gold dashed) is the mean of that tail — the coherent alternative.",
       x = "Daily Change (%)", y = "Density",
       caption = "VaR ignores the severity of tail losses. CVaR (Expected Shortfall) captures the average loss conditional on breaching VaR.") +
  theme_economist_quant()
```

# Section 8: Item 5 — Are the Differences Statistically Significant?

```{r item5_tables}
full_pct <- df_returns$dod_pct

item5_tests <- bind_rows(lapply(windows_returns[-1], function(w) {
  ks     <- ks.test(full_pct, w$dod_pct)
  levene <- levene_test_manual(full_pct, w$dod_pct)
  data.frame(
    n          = nrow(w),
    KS_Stat    = ks$statistic,
    KS_p       = ks$p.value,
    Levene_F   = abs(levene$statistic),
    Levene_p   = levene$p_value
  )
}), .id = "Window") %>%
  mutate(Window = factor(Window, levels = c("w60", "w40", "w20"))) %>%
  arrange(Window)

item5_desc <- bind_rows(lapply(windows_returns, function(w) {
  data.frame(n       = nrow(w),
             Mean    = mean(w$dod_pct),
             SD      = sd(w$dod_pct),
             Skew    = skewness(w$dod_pct),
             Kurt    = kurtosis(w$dod_pct))
}), .id = "Window") %>%
  mutate(Window = factor(Window, levels = window_order)) %>%
  arrange(Window)

item5_comp <- item1_var %>%
  mutate(Diff_from_Full_pp = VaR_Pct - VaR_Pct[Window == "Full"]) %>%
  arrange(Window)

kable(item5_tests, digits = 4,
      caption = "Item 5: Significance Tests (each window vs. full history). KS = distribution equality; Levene = variance equality.")
kable(item5_desc, digits = 4,
      caption = "Item 5: Descriptive Statistics by Window")
kable(item5_comp, digits = 4,
      caption = "Item 5: VaR Comparison — difference from full-history baseline (percentage points)")
```

```{r item5_var_bar}
# --- PLOT 4a: Stability in Numbers — VaR bar chart (ordered Full → w60 → w40 → w20) ---
ggplot(item1_var, aes(x = Window, y = VaR_Pct,
                      fill = Window == "Full")) +
  geom_col(width = 0.55) +
  scale_fill_manual(values = c("TRUE" = econ_blue, "FALSE" = econ_grey)) +
  geom_text(aes(label = paste0(round(VaR_Pct, 3), "%")),
            vjust = 1.6, color = "white", fontface = "bold", size = 4) +
  labs(title = "Stability in numbers",
       subtitle = "VaR 95% loss magnitude (%) by analysis window. Blue = full-history baseline (2019–2026).",
       x = NULL, y = "VaR Magnitude (%)",
       caption = paste0("Windows: Full = ", format(nrow(df_returns), big.mark=","), 
                        " obs; w60 = 60 obs; w40 = 40 obs; w20 = 20 obs.")) +
  theme_economist_quant() + theme(legend.position = "none")
```

```{r item5_dist_overlay}
# --- PLOT 4b: Distribution comparison — window return distributions overlaid ---
dist_data <- bind_rows(
  mutate(windows_returns$Full, Window = "Full (2019–2026)"),
  mutate(windows_returns$w60,  Window = "60-day"),
  mutate(windows_returns$w40,  Window = "40-day"),
  mutate(windows_returns$w20,  Window = "20-day")
) %>%
  mutate(Window = factor(Window,
                         levels = c("Full (2019–2026)", "60-day", "40-day", "20-day")))

win_colors <- c("Full (2019–2026)" = econ_blue,
                "60-day"           = econ_gold,
                "40-day"           = econ_grey,
                "20-day"           = econ_red)

ggplot(dist_data, aes(x = dod_pct, color = Window, fill = Window)) +
  geom_density(alpha = 0.12, linewidth = 0.7) +
  scale_color_manual(values = win_colors) +
  scale_fill_manual(values  = win_colors) +
  coord_cartesian(xlim = c(-2.5, 2.5)) +
  labs(title = "Same currency, different stories",
       subtitle = "Return distribution by analysis window. Short windows capture the current regime — not the full risk picture.",
       x = "Daily Close-to-Close Change (%)", y = "Density",
       caption = "KS and Levene tests assess whether differences from the full history are statistically significant.") +
  theme_economist_quant()
```

```{r item5_rolling_var}
# --- PLOT 4c: Rolling 60-day VaR over time ---
roll_n <- 60
roll_idx <- roll_n:nrow(df_returns)

roll_var_pct <- sapply(roll_idx, function(i) {
  win <- df_returns$dod_pct[(i - roll_n + 1):i]
  abs(quantile(win, 0.05, type = 7))
})

roll_df <- data.frame(
  Date    = df_returns$Date[roll_idx],
  RollVaR = roll_var_pct
)

# Mark current windows
current_w60_start <- min(windows_returns$w60$Date)

ggplot(roll_df, aes(x = Date, y = RollVaR)) +
  geom_line(color = econ_blue, linewidth = 0.6) +
  geom_hline(yintercept = item1_var$VaR_Pct[item1_var$Window == "Full"],
             linetype = "dashed", color = econ_grey, linewidth = 0.6) +
  annotate("text", x = as.Date("2019-06-01"),
           y = item1_var$VaR_Pct[item1_var$Window == "Full"] + 0.02,
           label = paste0("Full-history VaR: ",
                          round(item1_var$VaR_Pct[item1_var$Window == "Full"], 3), "%"),
           size = 3, color = econ_grey, hjust = 0) +
  geom_vline(xintercept = current_w60_start, color = econ_red,
             linetype = "dotted", linewidth = 0.7) +
  annotate("text", x = current_w60_start, y = max(roll_var_pct) * 0.97,
           label = "← 60-day window starts here",
           size = 3, color = econ_red, hjust = -0.05) +
  labs(title = "VaR is not static",
       subtitle = "Rolling 60-day VaR 95% (%) through time. Flat dashed line = full-history VaR benchmark.",
       x = NULL, y = "Rolling VaR 95% (%)",
       caption = "Demonstrates regime-dependence: short-window VaR reflects the current low-volatility period, not the full risk spectrum.") +
  theme_economist_quant()
```

# Section 9: Item 6 — Portfolio Loss and Hedge Sizing (99% Confidence)

```{r item6_calc}
# Reference: last trading day on or before 30 June 2024
ref_row      <- df %>% filter(Date <= as.Date("2024-06-30")) %>% slice_tail(n = 1)
fx_rate      <- ref_row$Close
ref_date_used <- ref_row$Date

# VaR and upside at 99% (1st and 99th percentile of full DoD return series)
var_99_pct    <- abs(quantile(df_returns$dod_pct, 0.01, type = 7))
upside_99_pct <-    quantile(df_returns$dod_pct, 0.99, type = 7)

# Position calculations
pos_usd           <- 10000000
loss_usd          <- (var_99_pct / 100) * pos_usd
loss_php          <- loss_usd * fx_rate

target_php        <- 20000
reduced_pos_usd   <- target_php / ((var_99_pct / 100) * fx_rate)
usd_to_remove     <- pos_usd - reduced_pos_usd

upside_full_php     <- (upside_99_pct / 100) * pos_usd * fx_rate
upside_reduced_php  <- (upside_99_pct / 100) * reduced_pos_usd * fx_rate
upside_sacrificed   <- upside_full_php - upside_reduced_php

item6_table <- data.frame(
  Parameter = c(
    "Reference Date Used", "FX Rate at Reference (PHP/USD)",
    "Full-History VaR 99% (%)", "VaR 99% (cents)",
    "USD Position", "1-Day Loss @ 99% (USD)", "1-Day Loss @ 99% (PHP)",
    "Target Maximum 1-Day Loss (PHP)", "Required USD Position",
    "USD to Redeploy", "99th Percentile Upside (%)",
    "Upside on Full Position (PHP)", "Upside on Reduced Position (PHP)",
    "Upside Sacrificed (PHP)"
  ),
  Value = c(
    as.character(ref_date_used),
    round(fx_rate, 4),
    round(var_99_pct, 4),
    round(abs(quantile(df_returns$dod_cents, 0.01, type = 7)), 4),
    format(pos_usd, big.mark = ","),
    format(round(loss_usd, 2), big.mark = ","),
    format(round(loss_php, 2), big.mark = ","),
    format(target_php, big.mark = ","),
    format(round(reduced_pos_usd, 2), big.mark = ","),
    format(round(usd_to_remove, 2), big.mark = ","),
    round(upside_99_pct, 4),
    format(round(upside_full_php, 2), big.mark = ","),
    format(round(upside_reduced_php, 2), big.mark = ","),
    format(round(upside_sacrificed, 2), big.mark = ",")
  )
)
kable(item6_table, caption = "Item 6: Portfolio Loss and Hedge Sizing (FX Rate as of 30 Jun 2024, 99% Confidence)")
```

```{r item6_viz}
# --- PLOT 5: The Price of Safety ---
port_viz <- data.frame(
  Category = factor(
    c("Full Exposure\n(USD 10M)", "Capped Loss\n(PHP 20K target)"),
    levels = c("Full Exposure\n(USD 10M)", "Capped Loss\n(PHP 20K target)")
  ),
  Loss   = c(loss_php / 1e6, target_php / 1e6),
  Upside = c(upside_full_php / 1e6, upside_reduced_php / 1e6)
) %>%
  pivot_longer(cols = c(Loss, Upside), names_to = "Direction", values_to = "Value") %>%
  mutate(
    SignedValue = ifelse(Direction == "Loss", -Value, Value),
    Direction   = factor(Direction, levels = c("Upside", "Loss"))
  )

ggplot(port_viz, aes(x = Category, y = SignedValue, fill = Direction)) +
  geom_col(width = 0.5, position = "identity") +
  geom_hline(yintercept = 0, color = "#333333", linewidth = 0.4) +
  scale_fill_manual(values = c("Loss" = econ_red, "Upside" = econ_blue),
                    labels = c("1-Day Loss (99%)", "99th-Pct Upside")) +
  scale_y_continuous(labels = function(x) paste0("PHP ", abs(x), "M"),
                     breaks = seq(-5, 5, 1)) +
  labs(title = "The price of safety",
       subtitle = "Potential 1-day loss (down) vs. upside potential (up), in PHP millions. Reference: FX rate as of 30 Jun 2024.",
       x = NULL, y = "PHP (millions)",
       caption = paste0("To cap daily loss at PHP 20,000, USD position must be reduced to USD ",
                        format(round(reduced_pos_usd, 0), big.mark = ","),
                        ". This sacrifices PHP ",
                        format(round(upside_sacrificed / 1e6, 2), big.mark = ","), "M in 99th-pct upside.")) +
  theme_economist_quant()
```

# Section 10: Summary Table

```{r summary}
summary_table <- data.frame(
  Item = c(1, 1, 1, 1, 1, 1, 2, 2, 2, 2, 2, 2, 3, 4, 4, 5, 5, 6, 6, 6),
  Description = c(
    "DoD VaR 95% — Full history",
    "DoD VaR 95% — 60-day window",
    "DoD VaR 95% — 40-day window",
    "DoD VaR 95% — 20-day window",
    "Outlier count (DoD, outside 0.1%/99.9%)",
    "Outlier dates (DoD)",
    "Intraday VaR 95% — Full history",
    "Intraday VaR 95% — 60-day window",
    "Intraday VaR 95% — 40-day window",
    "Intraday VaR 95% — 20-day window",
    "Outlier count (Intraday, above 99.9%)",
    "Outlier dates (Intraday)",
    "VaR gap (Intraday − DoD), full history",
    "CVaR / Expected Shortfall 95%, full",
    "VaR after Winsorization 95%",
    "KS test p-values (w60, w40, w20 vs Full)",
    "Max VaR deviation from full baseline",
    "1-day loss, USD 10M @ 99% (PHP)",
    "USD position to cap loss at PHP 20K",
    "99th-pct upside sacrificed (PHP)"
  ),
  Result = c(
    paste0(round(item1_var$VaR_Cents[item1_var$Window == "Full"], 2), " cts / ", round(item1_var$VaR_Pct[item1_var$Window == "Full"], 4), "%"),
    paste0(round(item1_var$VaR_Cents[item1_var$Window == "w60"], 2), " cts / ", round(item1_var$VaR_Pct[item1_var$Window == "w60"], 4), "%"),
    paste0(round(item1_var$VaR_Cents[item1_var$Window == "w40"], 2), " cts / ", round(item1_var$VaR_Pct[item1_var$Window == "w40"], 4), "%"),
    paste0(round(item1_var$VaR_Cents[item1_var$Window == "w20"], 2), " cts / ", round(item1_var$VaR_Pct[item1_var$Window == "w20"], 4), "%"),
    as.character(nrow(outliers_1)),
    paste(format(outliers_1$Date, "%d %b %Y"), collapse = "; "),
    paste0(round(item2_var$VaR_Cents[item2_var$Window == "Full"], 2), " cts / ", round(item2_var$VaR_Pct[item2_var$Window == "Full"], 4), "%"),
    paste0(round(item2_var$VaR_Cents[item2_var$Window == "w60"], 2), " cts / ", round(item2_var$VaR_Pct[item2_var$Window == "w60"], 4), "%"),
    paste0(round(item2_var$VaR_Cents[item2_var$Window == "w40"], 2), " cts / ", round(item2_var$VaR_Pct[item2_var$Window == "w40"], 4), "%"),
    paste0(round(item2_var$VaR_Cents[item2_var$Window == "w20"], 2), " cts / ", round(item2_var$VaR_Pct[item2_var$Window == "w20"], 4), "%"),
    as.character(nrow(outliers_2)),
    paste(format(outliers_2$Date, "%d %b %Y"), collapse = "; "),
    paste0(round(gap_table$Gap_Cents[gap_table$Window == "Full"], 2), " cts"),
    paste0(round(cvar_cents, 2), " cts / ", round(cvar_pct, 4), "%"),
    paste0(round(var_winsor_pct, 4), "% (reduction: ", round(raw_var_pct - var_winsor_pct, 4), " pp)"),
    paste(round(item5_tests$KS_p, 4), collapse = ", "),
    paste0(round(max(abs(item5_comp$Diff_from_Full_pp)), 4), " pp"),
    paste0("PHP ", format(round(loss_php, 2), big.mark = ",")),
    paste0("USD ", format(round(reduced_pos_usd, 2), big.mark = ",")),
    paste0("PHP ", format(round(upside_sacrificed, 2), big.mark = ","))
  )
)
kable(summary_table, caption = "Final Summary Table — All Items")
```

# Section 11: Strategic Risk Recommendations

Based on quantitative analysis of USD/PHP BAP data (`r format(min(df$Date), "%d %b %Y")` – `r format(max(df$Date), "%d %b %Y")`):

### 1. Ditch VaR for CVaR
Value at Risk (VaR) is not a **coherent** risk measure — it fails subadditivity and ignores tail severity. Switch to **Conditional VaR (CVaR / Expected Shortfall)**. CVaR measures the *average* loss in the worst 5% of scenarios rather than just the threshold, and behaves correctly when positions are aggregated.

### 2. Handle Outliers: Two Options
- **Winsorization**: Cap extreme returns at the 0.1%/99.9% boundary rather than removing them. The data remains complete; outlier influence is bounded. VaR falls slightly after winsorization, which shows the impact is measurable but not dramatic.
- **Switch to CVaR**: CVaR already captures the entire tail, so the *specific* outliers matter less — what matters is the average loss across all tail observations.

### 3. Short Windows Miss Structural Risk
The rolling VaR chart (Plot 4c) makes it plain: using only 20 or 60 days of data captures the *current regime*, not the full risk spectrum. The w60 VaR is materially lower than the full-history VaR, which includes the 2022 PHP depreciation and the COVID-19 volatility spike. For prudent risk management, the full history is the appropriate baseline.

### 4. Trade Dollars for Pesos — But Know the Cost
To cap daily losses at PHP 20,000, capital must move from USD into **PHP cash or T-bills**. This exploits **Translation Invariance** of coherent risk measures: $\rho(X - c) = \rho(X) - c$, where adding a risk-free asset of the base currency reduces total risk by exactly that amount. However, capping the loss at PHP 20,000 requires gutting the USD position so severely that approximately **PHP `r format(round(upside_sacrificed / 1e6, 2))`M** in 99th-percentile upside is sacrificed. Risk reduction is never free.

---
