---
title: "USD/PHP FX Risk Analysis | Items 1-6"
output: 
  html_document:
    df_print: kable
    toc: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 6)
library(readr)
library(dplyr)
library(lubridate)
library(knitr)
library(e1071)
library(ggplot2)
library(ggrepel)

# --- ECONOMIST & QVRS THEME ---
theme_economist_quant <- function() {
  theme_minimal() +
    theme(
      text = element_text(family = "sans"),
      plot.title = element_text(face = "bold", size = 16, color = "#0D0D0D"),
      plot.subtitle = element_text(size = 11, color = "#595959", margin = margin(b = 10)),
      plot.caption = element_text(size = 8, color = "#B3B3B3", hjust = 0, margin = margin(t = 10)),
      panel.grid.major.y = element_line(color = "#EBEBEB", linewidth = 0.3),
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      axis.title = element_text(size = 10, face = "bold", color = "#333333"),
      axis.ticks = element_blank(),
      legend.position = "bottom",
      legend.title = element_blank()
    )
}

econ_blue <- "#141F52" # Chicago 20
econ_red  <- "#E3120B" # Economist Red
econ_grey <- "#595959" # London 35

# Helper for Levene's Test (Manual Implementation)
levene_test_manual <- function(group1, group2) {
  med1 <- median(group1)
  med2 <- median(group2)
  z1 <- abs(group1 - med1)
  z2 <- abs(group2 - med2)
  t_res <- t.test(z1, z2, var.equal = TRUE)
  return(list(statistic = t_res$statistic, p_value = t_res$p.value))
}
```

# Section 1: Data Load and Integrity Check

```{r data_load}
df_raw <- read_csv("data/unified_usdphp_bap.csv") %>%
  mutate(Date = as.Date(Date)) %>%
  # Filter for trading days only (Monday=2, ..., Friday=6)
  # This removes the weekends imputed by the consolidation script
  filter(wday(Date) %in% 2:6) %>%
  # Fix specific outlier on 2021-05-20: 478.855 -> 47.855
  mutate(Open = ifelse(Date == as.Date("2021-05-20") & Open == 478.855, 47.855, Open)) %>%
  arrange(Date)

integrity_check <- data.frame(
  Metric = c("Total Trading Row Count", "Start Date", "End Date", "Rows with High < Low"),
  Value = c(
    as.character(nrow(df_raw)),
    as.character(min(df_raw$Date)),
    as.character(max(df_raw$Date)),
    as.character(nrow(filter(df_raw, High < Low)))
  )
)
kable(integrity_check, caption = "Data Integrity Check")
```

# Section 2: Data Cleaning

```{r data_cleaning}
# Swap High/Low errors
df <- df_raw %>%
  mutate(
    ActualHigh = pmax(High, Low),
    ActualLow = pmin(High, Low),
    High = ActualHigh,
    Low = ActualLow
  ) %>%
  select(-ActualHigh, -ActualLow)

# Compute Derived Columns
df <- df %>%
  mutate(
    dod_cents = (Close - lag(Close)) * 100,
    dod_pct = (Close - lag(Close)) / lag(Close) * 100,
    intraday_cents = (High - Low) * 100,
    intraday_pct = (High - Low) / Low * 100
  )

# Separate for DoD analysis (drop NA row)
df_returns <- df %>% filter(!is.na(dod_cents))

# Cleaning Verification Output
cleaning_stats <- data.frame(
  Metric = c("Swapped High/Low Count", "Derived Columns", "Remaining Trading Rows"),
  Result = c(
    as.character(nrow(filter(df_raw, High < Low))),
    "DoD (cents/%), Intraday (cents/%)",
    as.character(nrow(df_returns))
  )
)
kable(cleaning_stats, caption = "Data Cleaning Results Summary")
```


# Section 3: Window Definitions

```{r windows}
# Applied to return series
windows_returns <- list(
  Full = df_returns,
  w60 = tail(df_returns, 60),
  w40 = tail(df_returns, 40),
  w20 = tail(df_returns, 20)
)

# Applied to intraday series
windows_intraday <- list(
  Full = df,
  w60 = tail(df, 60),
  w40 = tail(df, 40),
  w20 = tail(df, 20)
)

window_summary <- data.frame(
  Window = names(windows_returns),
  Start_Date = sapply(windows_returns, function(x) as.character(min(x$Date))),
  n = sapply(windows_returns, nrow)
)
kable(window_summary, caption = "Analysis Windows (Return Series)")
```

# Section 4: Item 1 Tables (DoD VaR 95%)

```{r item1_tables}
item1_var <- bind_rows(lapply(windows_returns, function(w) {
  data.frame(
    n = nrow(w),
    # VaR as absolute loss magnitude
    VaR_Pct = abs(quantile(w$dod_pct, 0.05, type = 7)),
    VaR_Cents = abs(quantile(w$dod_cents, 0.05, type = 7))
  )
}), .id = "Window")
kable(item1_var, caption = "Item 1: Close-to-Close VaR at 95% (Loss Magnitude)")

# Outliers
lb_999 <- quantile(df_returns$dod_pct, 0.001, type = 7)
ub_999 <- quantile(df_returns$dod_pct, 0.999, type = 7)
outliers_1 <- df_returns %>% 
  filter(dod_pct < lb_999 | dod_pct > ub_999) %>%
  select(Date, Close, dod_pct, dod_cents)
kable(outliers_1, caption = "Item 1: Outliers (0.1% / 99.9% Bounds)")
```

```{r item1_viz}
# --- PLOT 1: The Long Tail of Risk ---
ggplot(df_returns, aes(x = dod_pct)) +
  stat_function(fun = dnorm, args = list(mean = mean(df_returns$dod_pct), sd = sd(df_returns$dod_pct)), 
                linetype = "dashed", color = econ_grey, linewidth = 0.5) +
  geom_density(fill = econ_blue, alpha = 0.2, color = econ_blue) +
  geom_vline(xintercept = -item1_var$VaR_Pct[1], color = econ_blue, linewidth = 1) +
  geom_text(aes(x = -item1_var$VaR_Pct[1], y = 0.2, label = paste("VaR 95%:", round(item1_var$VaR_Pct[1], 3), "%")), 
            angle = 90, vjust = -1, size = 3, color = econ_blue) +
  labs(title = "The long tail of risk",
       subtitle = "Density of USD/PHP daily percentage changes vs. Normal Distribution (grey dashed), 2019-2026",
       x = "Daily Change (%)", y = "Density",
       caption = "Source: BAP Historical Data. VaR calculated using historical simulation (type 7).") +
  theme_economist_quant()
```

# Section 5: Item 2 Tables (Intraday VaR 95%)

```{r item2}
item2_var <- bind_rows(lapply(windows_intraday, function(w) {
  data.frame(
    n = nrow(w),
    VaR_Cents = quantile(w$intraday_cents, 0.95, type = 7),
    VaR_Pct = quantile(w$intraday_pct, 0.95, type = 7)
  )
}), .id = "Window")
kable(item2_var, caption = "Item 2: Intraday VaR at 95% (95th percentile of Range)")

# Outliers
ub_intraday <- quantile(df$intraday_cents, 0.999, type = 7)
outliers_2 <- df %>% 
  filter(intraday_cents > ub_intraday) %>%
  select(Date, High, Low, intraday_cents, intraday_pct)
kable(outliers_2, caption = "Item 2: Outliers (99.9% Bound)")
```

# Section 6: Item 3 Table (Gap Analysis)

```{r item3_table}
gap_table <- item1_var %>%
  select(Window, dod_var = VaR_Cents) %>%
  left_join(item2_var %>% select(Window, intra_var = VaR_Cents), by = "Window") %>%
  mutate(Gap = abs(intra_var) - abs(dod_var))
kable(gap_table, caption = "Item 3: VaR Gap (Intraday - DoD)")
```

```{r item3_viz}
# --- PLOT 2: Mind the Gap ---
# Filter last 250 days for visibility
df_recent <- tail(df, 250)

ggplot(df_recent, aes(x = Date)) +
  geom_line(aes(y = intraday_pct, color = "Intraday Range"), linewidth = 0.5) +
  geom_line(aes(y = abs(dod_pct), color = "Absolute DoD Change"), linewidth = 0.5) +
  scale_color_manual(values = c("Intraday Range" = econ_blue, "Absolute DoD Change" = econ_red)) +
  labs(title = "Mind the gap",
       subtitle = "Intraday volatility (%) vs. Absolute Close-to-Close changes (%), trailing 250 days",
       x = NULL, y = "Magnitude (%)",
       caption = "Large gaps indicate intraday chaos that 'recovers' by the closing bell.") +
  theme_economist_quant()
```

# Section 7: Item 4 Table (CVaR & Winsorization)

```{r item4}
# CVaR (Expected Shortfall)
# Threshold is the 5th percentile return
threshold_pct <- -item1_var$VaR_Pct[1]
threshold_cents <- -item1_var$VaR_Cents[1]

cvar_pct <- abs(mean(df_returns$dod_pct[df_returns$dod_pct <= threshold_pct]))
cvar_cents <- abs(mean(df_returns$dod_cents[df_returns$dod_cents <= threshold_cents]))

# Winsorization
lb_999 <- quantile(df_returns$dod_pct, 0.001, type = 7)
ub_999 <- quantile(df_returns$dod_pct, 0.999, type = 7)

df_winsor <- df_returns %>%
  mutate(dod_pct_w = pmin(pmax(dod_pct, lb_999), ub_999),
         dod_cents_w = pmin(pmax(dod_cents, quantile(dod_cents, 0.001, type=7)), quantile(dod_cents, 0.999, type=7)))

var_winsor_pct <- abs(quantile(df_winsor$dod_pct_w, 0.05, type = 7))
var_winsor_cents <- abs(quantile(df_winsor$dod_cents_w, 0.05, type = 7))

item4_table <- data.frame(
  Metric = c("Raw VaR (95%)", "CVaR (Expected Shortfall)", "Winsorization LB (0.1%)", "Winsorization UB (99.9%)", "VaR after Winsorization", "VaR reduction"),
  Percentage = c(item1_var$VaR_Pct[1], cvar_pct, lb_999, ub_999, var_winsor_pct, item1_var$VaR_Pct[1] - var_winsor_pct),
  Cents = c(item1_var$VaR_Cents[1], cvar_cents, NA, NA, var_winsor_cents, item1_var$VaR_Cents[1] - var_winsor_cents)
)
kable(item4_table, caption = "Item 4: Results (Full History)")
```

# Section 8: Item 5 Tables (Significance)

```{r item5_tables}
full_pct <- df_returns$dod_pct
item5_tests <- bind_rows(lapply(windows_returns[-1], function(w) {
  ks <- ks.test(full_pct, w$dod_pct)
  levene <- levene_test_manual(full_pct, w$dod_pct)
  data.frame(n = nrow(w), KS_Stat = ks$statistic, KS_p = ks$p.value, Levene_Stat = abs(levene$statistic), Levene_p = levene$p_value)
}), .id = "Window")

item5_desc <- bind_rows(lapply(windows_returns, function(w) {
  data.frame(n = nrow(w), Mean_Pct = mean(w$dod_pct), SD_Pct = sd(w$dod_pct), Skewness = skewness(w$dod_pct), Kurtosis = kurtosis(w$dod_pct))
}), .id = "Window")

item5_comp <- item1_var %>%
  mutate(Diff_from_Full = VaR_Pct - VaR_Pct[1])

kable(item5_tests, caption = "Item 5: Significance Tests")
kable(item5_desc, caption = "Item 5: Descriptive Statistics")
kable(item5_comp, caption = "Item 5: VaR Comparison")
```

```{r item5_viz}
# --- PLOT 3: Stability in Numbers ---
ggplot(item1_var, aes(x = Window, y = VaR_Pct, fill = Window == "Full")) +
  geom_col(width = 0.6) +
  scale_fill_manual(values = c("TRUE" = econ_blue, "FALSE" = econ_grey)) +
  geom_text(aes(label = paste(round(VaR_Pct, 3), "%")), vjust = 1.5, color = "white", fontface = "bold") +
  labs(title = "Stability in numbers",
       subtitle = "VaR (95%) loss magnitude across windows. Differences from historical are statistically examined.",
       x = NULL, y = "VaR Magnitude (%)",
       caption = "Blue bar represents historical baseline since 2019.") +
  theme_economist_quant() + theme(legend.position = "none")
```

# Section 9: Item 6 Table (Portfolio)

```{r item6_calc}
ref_row <- df %>% filter(Date <= as.Date("2024-06-30")) %>% slice_tail(n = 1)
fx_rate <- ref_row$Close
ref_date_used <- ref_row$Date

# VaR at 99% level of confidence (1st percentile return)
var_99_pct <- abs(quantile(df_returns$dod_pct, 0.01, type = 7))
upside_99_pct <- quantile(df_returns$dod_pct, 0.99, type = 7)

pos_usd <- 10000000
loss_usd <- (var_99_pct / 100) * pos_usd
loss_php <- loss_usd * fx_rate

target_php <- 20000
reduced_pos_usd <- target_php / ((var_99_pct / 100) * fx_rate)
upside_full_php <- (upside_99_pct / 100) * pos_usd * fx_rate
upside_reduced_php <- (upside_99_pct / 100) * reduced_pos_usd * fx_rate
upside_sacrificed_php <- upside_full_php - upside_reduced_php

item6_table <- data.frame(
  Parameter = c("Reference Date Used", "FX Rate (PHP/USD)", "VaR at 99% (%)", "USD Position", "1-Day Loss (USD)", "1-Day Loss (PHP)", "Target Max Loss (PHP)", "Reduced USD Position", "USD to Remove", "Upside Return 99% (%)", "Upside Full (PHP)", "Upside Reduced (PHP)", "Upside Sacrificed (PHP)"),
  Value = c(as.character(ref_date_used), fx_rate, var_99_pct, pos_usd, loss_usd, loss_php, target_php, reduced_pos_usd, pos_usd - reduced_pos_usd, upside_99_pct, upside_full_php, upside_reduced_php, upside_sacrificed_php)
)
kable(item6_table, caption = "Item 6: Portfolio Loss and Hedge Sizing")
```

```{r item6_viz}
# --- PLOT 4: The Price of Safety ---
port_viz <- data.frame(
  Category = c("Full Exposure\n(PHP 10M)", "Capped Loss\n(Target)"),
  Loss = c(loss_php / 1e6, target_php / 1e6),
  Upside = c(upside_full_php / 1e6, upside_reduced_php / 1e6)
)

ggplot(port_viz, aes(x = Category)) +
  geom_col(aes(y = -Loss, fill = "1-Day Loss (99%)"), width = 0.5) +
  geom_col(aes(y = Upside, fill = "99th Percentile Upside"), width = 0.5) +
  scale_fill_manual(values = c("1-Day Loss (99%)" = econ_red, "99th Percentile Upside" = econ_blue)) +
  labs(title = "The price of safety",
       subtitle = "Potential 1-day loss vs. upside potential in millions of PHP (Reference: June 2024)",
       x = NULL, y = "PHP (Millions)",
       caption = "Capping loss at PHP 20,000 virtually eliminates potential upside potential.") +
  theme_economist_quant()
```

# Section 10: Summary Table

```{r summary}
summary_table <- data.frame(
  Item = c(1, 1, 2, 2, 3, 4, 4, 5, 5, 6, 6, 6),
  Description = c("Close-to-close VaR 95% full", "Outlier count (DoD)", "Intraday VaR 95% full", "Outlier count (Intraday)", "VaR gap full", "CVaR 95% full", "Winsorization adjustment", "KS test p-values (60,40,20)", "Max VaR diff across windows", "1-day loss USD 10M (PHP)", "USD pos to cap at PHP 20K", "Upside sacrificed"),
  Result = c(
    paste(round(item1_var$VaR_Cents[1], 2), "cents /", round(item1_var$VaR_Pct[1], 4), "%"),
    as.character(nrow(outliers_1)),
    paste(round(item2_var$VaR_Cents[1], 2), "cents /", round(item2_var$VaR_Pct[1], 4), "%"),
    as.character(nrow(outliers_2)),
    paste(round(gap_table$Gap[1], 2), "cents"),
    paste(round(cvar_cents, 2), "cents /", round(cvar_pct, 4), "%"),
    paste(round(item1_var$VaR_Pct[1] - var_winsor_pct, 4), "pp"),
    paste(round(item5_tests$KS_p, 4), collapse = ", "),
    paste(round(max(abs(item5_comp$Diff_from_Full)), 4), "pp"),
    paste("PHP", format(round(loss_php, 2), big.mark=",")),
    paste("USD", format(round(reduced_pos_usd, 2), big.mark=",")),
    paste("PHP", format(round(upside_sacrificed_php, 2), big.mark=","))
  )
)
kable(summary_table, caption = "Final Summary Table")
```

# Section 11: Strategic Risk Recommendations

Based on the quantitative analysis of USD/PHP data (2019-2026), these four steps will improve the firm's risk posture:

### 1. Ditch VaR for CVaR
Value at Risk (VaR) is a flawed metric. It is not "coherent" because it fails the test of subadditivity: it can wrongly suggest that a diversified portfolio is riskier than the sum of its parts. The institution should switch to **Conditional Value at Risk (CVaR)**. Unlike VaR, which ignores the "tail," CVaR accounts for the worst-case scenarios and behaves logically when assets are combined.

### 2. Trade Dollars for Pesos
To cap daily losses at PHP 20,000, move capital out of dollars and into **PHP cash or T-bills**. This exploits the **Translation Invariance** of coherent risk measures: moving assets into the base currency reduces total risk $\rho(X)$ by exactly the amount moved. It is a mathematical certainty, not a forecast.

### 3. Face the Price of Safety
Risk management is never a free lunch. Capping one-day losses at PHP 20,000 requires gutting the dollar position so severely that it eliminates potential gains. This safety comes at an opportunity cost of approximately **PHP 5.4m** in 99th-percentile upside. Managers must decide if avoiding a small daily loss is worth sacrificing millions in potential profit.

---

# Appendix: Data Integrity Findings

**The `unified_usdphp_bap.csv` file is used as the primary data source for this analysis.**

## Verification Results

A robust heuristic check across all eight Excel files accounted for their inconsistent structures.

```{r data_integrity_table, echo=FALSE}
integrity_summary <- data.frame(
  Source = c("Unified CSV (Trading Days)", "Excel Files (Total)", "Difference"),
  Unique_Dates = c("~1,800", "1,486", "+~314"),
  Date_Range = c("2019-01-02 to 2026-02-27", "2019-01-02 to 2025-02-18", "Missing Mar 2025 – Feb 2026")
)
kable(integrity_summary, col.names = c("Source", "Unique Dates", "Date Range"), caption = "Data Source Comparison")
```

